### 2020-10-02 manuscript/paper_BRM.Rmd
Fix minor issues

abstract: |
  Eye-tracking allows researchers to infer cognitive processes from eye movements that are classified into distinct events. Parsing the events is typically done by algorithms. Previous algorithms have successfully used hidden Markov models (HMMs) for classification but still inhere several weaknesses. To adress these weaknesses, we developed gazeHMM, an algorithm that uses an HMM as a generative model, has no critical parameters to be set by users, and does not require human coded data as input. The algorithm classifies gaze data into fixations, saccades, and optionally PSOs and smooth pursuits. We evaluated the performance of gazeHMM in a simulation study and on two benchmark data sets. The simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences. Critical exceptions for good recovery were including the smooth pursuit state in the generative model and noisy data. The performance on benchmark data was assessed by model comparison and agreement of the classification to human coding compared to other algorithms. The model comparison yielded preferred generative models for the benchmark data with more events than we expected. Nevertheless, gazeHMM showed relatively high similarity to human coding and outperformed the other algorithms in some agreement criteria. For dynamic stimuli, our algorithm tended to switch rapidly between fixations and smooth pursuits. Thus, we recommend to use gazeHMM for classifiying smooth pursuits only in explorative settings. Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to improve the classification of smooth pursuits.

@@

Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017]. Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015]. Basic events are fixations, saccades, smooth pursuits, and post-saccadic oscillations (PSOs). Classifying raw eye-tracking data into these events reduces their complexity and is usually the first step towards cognitive interpretation [@Salvucci2000]. The classification is typically done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017]. @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].  
On one hand, many of these classification algorithms use non-parametric methods to differentiate between eye movement events. For instance, the classic "Velocity-threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stampe1993]. On the other hand, many parametric methods have been developed recently. Some of them require human labeled training data as input and can therefore be termed as supervised. For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders. However, these supervised methods have several disadvantages: First, they require a sufficient amount of labeled data (usually labeled by humans) as input, which can become costly and time consuming [@Zemblys2019]. Second, they often rely on hyperparameters which are difficult to theoretically justify. Third, most supervised methods operate in a black-box fashion, which implies that they do not make any explicit assumptions about the data generating process [cf. @Zembyls2019]. Beyond that, there are several algorithms using unsupervised techniques. They typically do not need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers [e.g., @Houpt2018; @Hessels2017].  
Besides discriminating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data. To our knowledge, these *generative models* have been rarely used to classify eye movement data [cf. @Mihali2017; @Wadehn2019]. However, they come with various advantages and we will summarize the two most important ones for eye movement classification here [for an overview, see @Epstein2008].  
First, generative models allow to not only predict (classify) but also explain eye-tracking phenomena. For instance, they can inhere parameters that explain why oscillations only occur after but not before saccades. When the goal is to understand eye movement events and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods. Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions [@Epstein2008]. For example, they might suggest that oscillations also occur before saccadic eye movements [as has been mentioned by @Nystrom2010] or that the assumption of discrete eye movements does not hold [as has been discussed in @Andersson2017].  

@@

In sum, HMMs seem to be a promising method for classifying eye movements. However, the existing HMM algorithms each have their weaknesses that could be improved upon.  

@@

For this article, we developed a novel algorithm for classifying gaze data, named gazeHMM, that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms. The development was guided by four goals: (a) the algorithm should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it should confirm rather than explore the presence of events in the data.  
The following section will describe gazeHMM and the underlying HMM in detail. Then, we will present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms with regard to the agreement to human coding. Finally, we will discuss these results and propose directions in which gazeHMM and other HMM algorithms could be improved.

@@

p1 <- ggplot(toPlot, aes(x = t, x.va)) + 
  geom_line(aes(y = x.va)) + 
  geom_line(aes(y = y.va)) + 
  geom_point(aes(y = x.va), color = "darkred", size = 0.9) + 
  geom_point(aes(y = y.va), color = "darkblue", size = 0.9) +
  scale_y_continuous(name = "Gaze position (deg)", limits = c(-5, 10), breaks = pretty(c(-5, 10)))

p2 <- ggplot(toPlot, aes(x = t, y = vel)) + 
  geom_line() + 
  geom_point(size = 0.9) +
  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))

p3 <- ggplot(toPlot, aes(x = t, y = vel)) + 
  geom_line() + 
  geom_point(aes(color = event.pre), size = 0.9) +
  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))

p4 <- ggplot(toPlot, aes(x = t, y = vel)) + 
  geom_line() + 
  geom_point(aes(color = event.post), size = 0.9) +
  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))

@@

## Preprocessing
Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events. Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017]. However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007]. This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's sampling frequency [@Andersson2017]. In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, we assume that the angle between samples should not follow any direction but a uniformly random walk [@Larsson2015]. In contrast, saccade samples usually have a high velocity and acceleration and roughly follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, these events were not in the scope of gazeHMM.  
The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010]. The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan\left(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}\right) -\arctan\left(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}\right),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$, and is therefore bound between 0 and $2\pi$.  
Most of the missing data in eye movement classification are due to blinks. Therefore, the user can indicate which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, gazeHMM removes them heuristically. Before classifying the samples, it sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.

@@

The parameters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997]. The EM algorithm is generally suitable to estimate likelihoods with missing variables. For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data and the expected hidden states [eye movement events; @Visser2011]. When evaluating the likelihood for missing data, gazeHMM integrates over all possible values, which results in a probability density of one. The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability. Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988]. The implementation in depmixS4 allows to include time varying covariates for each parameter in the HMM. In gazeHMM, no such covariates were included and thus, only intercepts were estimated for each parameter.

## Postprocessing
After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence. We implemented this routine because constraining the probabilities for nonsaccade events to turn into PSOs to zero often caused PSOs not to appear in the state sequence at all. Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events. Thus, the postprocessing routine heuristically compensates for such violations. This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow nonsaccade events. Samples are relabeled as the state of the previous event. Finally, samples initially indicated as missing are labeled as noise (including blinks) and event descriptives are computed (e.g., fixation duration).  
The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages \texttt{signal} [@Ligges2015] to compute velocity and acceleration signals, \texttt{depmixS4} [@Visser2010] for the HMM, \texttt{CircStats} [@Lund2018] for the von Mises distribution, and \texttt{BB} [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization. The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).

@@

## Results
### Parameter Variation
In the first part of the simulation, we examined how varying the parameters[^2] in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences. For the two-state HMM, the recovery of parameters and states was nearly perfect (all RMdSPDs < 0.1, intercepts and slopes of regression lines almost zero and one, respectively, and Cohen's kappa close to 1). We therefore chose to include the respective figures in the supplementary material.  
For the HMM with three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3). When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently between 0.1 and 0.5. Varying $\kappa$ in states two and three led to RMdSPDs between 0.1 and 0.5 in the respective states, which we interpreted as moderate recovery. Otherwise, RMdSPDs were consistently lower than 0.1, indicating good recovery. Inspecting the regression lines between true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed strong and unbiased linear relationships (intercepts close to zero and slopes close to one). In contrast to the two-state HMM, larger deviations and more outliers were observed. Cohen's kappa values are presented in Figure \@ref(fig:plot-acc-1-3). For most estimated models, the kappa values between true and estimated state sequences were above 0.95, meaning almost perfect agreement. However, for some models, we observed kappas clustered around zero or -0.33, which suggests that state labels were switched (see supplementary material).  
[^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$. Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$). Therefore, we decided not to include it in the analysis.

@@

Concerning overall disagreement, Figure \@ref(fig:plot-disag) shows that gazeHMM had less disagreement to the human coders across all events for all three conditions. For the static condition, the difference to most other algorithms was only slight ($\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Image"]][3] - A2017.disag[["Image"]][-c(1:4)]))`$%), but for the dynamic conditions, the differences were substantial (video: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Video"]][4] - A2017.disag[["Video"]][-c(1:4)]))`$%) and large (dots: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Dots"]][4] - A2017.disag[["Dots"]][-c(1:4)]))`$%).  
To explore which events gazeHMM classified differently than the average of human coder, we looked at the confusion matrix between the two (see Table \@ref(tab:tab-conf-4)). It can be seen that gazeHMM classified many fixations as smooth pursuits and vice versa. Moreover, it confused many PSOs with saccades. The heuristic to detect blinks seemed to work successfully since gazeHMM classified most blink samples in agreement with human coding and only a minor part was mistaken for saccades. Inspecting an example of gaze data classified by gazeHMM compared to human coding leads to a similar notion: Figure \@ref(fig:plot-example-classification) illustrates that gazeHMM is rapidly switching between classifying fixations and smooth pursuits, whereas the human coder treats the same data as one large smooth pursuit event. In the example, the human coder also delayed the start of the PSO compared to gazeHMM.

@@

In this report, we presented gazeHMM, a novel algorithm for classifying gaze data into eye movement events. The algorithm models velocity, acceleration, and sample-to-sample angle signals with gamma distributions and a mixture of von Mises and a uniform distribution. An HMM serves as the generative model of the algorithm and classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits. We showed in a simulation study that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions. Adding a fourth event (i.e., smooth pursuit) to the model and noise to the generated data were the most critical factors for decreasing recovery. Furthermore, we applied gazeHMM with different numbers of states to benchmark data by @Andersson2017 and compared the model fit. The model comparison revealed that a five-state HMM had consistently most likely generated the data. This result opposed our expectation that a three-state model would be preferred for static and a four-state model for dynamic data. When comparing gazeHMM against other algorithms, gazeHMM showed mostly good agreement to human coding. On one hand, it outperformed the other algorithms in the overall disagreement to human coding. On the other hand, gazeHMM confused a lot of fixations with smooth pursuits, which led to a rapid switching between the two events. It also tended to mistake PSO samples as belonging to saccades.  

@@

*The submodels of gazeHMM were misspecified*. @Pohle2017 identified two scenarios in which model recovery using the ICL did not give optimal results: Outliers in the data and inadequate distributions in the response models. Both situations could apply to gazeHMM and eye movement data. Outliers occur frequently in eye-tracking data due to measurement error. Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task from which the data are obtained from [@Langrock2015]. Moreover, gazeHMM only estimated intercepts for all parameters and thus, no time-varying covariates were included [cf. @Li2017]. This aspect could indeed oversimplify the complex nature of eye movement data.  

@@

## Conclusion
Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events. For smooth pursuits, the classification is not optimal and thus not recommended, yet. On one side, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled input being the most important one. On the other side, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events. These difficulties highlight challenges for HMMs in general but also specifically for eye movements.

>>>>

abstract: |
  Eye-tracking allows researchers to infer cognitive processes from eye movements that are classified into distinct events. Parsing the events is typically done by algorithms. Previous algorithms have successfully used hidden Markov models (HMMs) for classification but still inhere several weaknesses. To adress these weaknesses, we developed gazeHMM, an algorithm that uses an HMM as a generative model, has no critical parameters to be set by users, and does not require human coded data as input. The algorithm classifies gaze data into fixations, saccades, and optionally postsaccadic oscillations and smooth pursuits. We evaluated the performance of gazeHMM in a simulation study and on two benchmark data sets. The simulation study showed that gazeHMM successfully recovered HMM parameters and hidden state sequences. Critical exceptions for good recovery were including the smooth pursuit state in the generative model and noisy data. The performance on benchmark data was assessed by model comparison and agreement of the classification to human coding compared to other algorithms. The model comparison yielded preferred generative models for the benchmark data with more events than we expected. Nevertheless, gazeHMM showed relatively high similarity to human coding and outperformed the other algorithms in some agreement criteria. For dynamic stimuli, our algorithm tended to switch rapidly between fixations and smooth pursuits. Thus, we recommend using gazeHMM for classifiying smooth pursuits only in explorative settings. Future HMM algorithms could use covariates to better capture eye movement processes and explicitly model event durations to improve the classification of smooth pursuits.

@@

Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017]. Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015]. Basic events are fixations, saccades, smooth pursuits, and post-saccadic oscillations (PSOs). Classifying raw eye-tracking data into these events reduces their complexity and is usually the first step towards cognitive interpretation [@Salvucci2000]. The classification is typically done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017]. @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019]. To motivate our decision to add another algorithm to this already rich selection of classification tools, we first need to distinguish between the different classes of methods they use.   
On one hand, many classification algorithms use non-parametric methods to differentiate between eye movement events. For instance, the classic "Velocity-threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stampe1993]. On the other hand, many parametric methods have been developed recently. Some of them require human labeled training data as input and can therefore be termed as supervised. For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders. Despite their accuracy, supervised methods have several disadvantages: First, they require a sufficient amount of labeled data (usually labeled by humans) as input, which can become costly and time consuming [@Zemblys2019]. Second, they often rely on hyperparameters which are difficult to theoretically justify. Third, most supervised methods operate in a black-box fashion, which implies that they do not make any explicit assumptions about the data generating process [cf. @Zemblys2019]. Beyond that, there are several algorithms using unsupervised techniques. They typically do not need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers [e.g., @Houpt2018; @Hessels2017].  
Besides discriminating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data. To our knowledge, these *generative models* have been rarely used to classify eye movement data [cf. @Mihali2017; @Wadehn2020]. They come with various advantages and we will summarize the two most important ones for eye movement classification here [for an overview, see @Epstein2008].  
First, generative models allow to not only predict (classify) but also help to explain eye-tracking phenomena. For instance, they can inhere parameters that indicate that oscillations only occur after but not before saccades. When the goal is to understand eye movement events and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods. Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions [@Epstein2008]. For example, they might suggest that oscillations also occur before saccadic eye movements [as has been mentioned by @Nystrom2010] or that the assumption of discrete eye movements does not hold [as has been discussed in @Andersson2017].  

@@

In sum, HMMs seem to be a promising method for classifying eye movements. Nevertheless, the existing HMM algorithms each have their weaknesses that could be improved upon.  

@@

For this article, we developed a novel algorithm for classifying gaze data, named gazeHMM, that relies on an HMM as a generative model and addresses the weaknesses of previous algorithms. The development was guided by four goals: (a) the algorithm should not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it should cover the most relevant eye movement events, namely fixations, saccades, smooth pursuits, and PSOs; (d) it should confirm rather than explore the presence of events in the data.  
The following section will describe gazeHMM and the underlying generative model in detail. Then, we will present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms with regard to the agreement to human coding. Finally, we will discuss these results and propose directions in which gazeHMM and other HMM algorithms could be improved.

@@

p1 <- ggplot(toPlot, aes(x = t, x.va)) + 
  geom_line(aes(y = x.va)) + 
  geom_line(aes(y = y.va)) + 
  geom_point(aes(y = x.va), color = "darkred", size = 1) + 
  geom_point(aes(y = y.va), color = "darkblue", size = 1) +
  scale_y_continuous(name = "Gaze position (deg)", limits = c(-5, 10), breaks = pretty(c(-5, 10)))

p2 <- ggplot(toPlot, aes(x = t, y = vel)) + 
  geom_line() + 
  geom_point(size = 1) +
  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))

p3 <- ggplot(toPlot, aes(x = t, y = vel)) + 
  geom_line() + 
  geom_point(aes(color = event.pre), size = 1) +
  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))

p4 <- ggplot(toPlot, aes(x = t, y = vel)) + 
  geom_line() + 
  geom_point(aes(color = event.post), size = 1) +
  scale_y_continuous(name = "Velocity (deg/s)", limits = c(0, 250), breaks = pretty(c(0, 250)))

@@

## Preprocessing
Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events. Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017], but most of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007]. This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's sampling frequency [@Andersson2017]. In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.  
Theoretically, these three metrics should clearly separate eye movement events. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, we assume that the angle between samples should not follow any direction but a uniformly random walk [@Larsson2015]. In contrast, saccade samples usually have a high velocity and acceleration and roughly follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, these events were not in the scope of gazeHMM.  
The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010]. The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan\left(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}\right) -\arctan\left(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}\right),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$, and is therefore bound between 0 and $2\pi$.  
Most of the missing data in eye movement classification are due to blinks. Therefore, the user can indicate which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, gazeHMM removes them heuristically. Before classifying the samples, it sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.

@@

The parameters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997]. The EM algorithm is generally suitable to estimate likelihoods with missing variables. For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data and the expected hidden states [i.e., eye movement events; @Visser2011]. When evaluating the likelihood for missing data, gazeHMM integrates over all possible values, which results in a probability density of one. The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability. Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988]. The implementation in depmixS4 allows to include time varying covariates for each parameter in the HMM. In gazeHMM, no such covariates were included and thus, only intercepts were estimated for each parameter.

## Postprocessing
After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence. We implemented this routine because constraining the probabilities for nonsaccade events to turn into PSOs to zero often caused PSOs not to appear in the state sequence at all. Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events. Thus, the postprocessing routine heuristically compensates for such violations. This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold (here: 10 ms), and PSOs that follow nonsaccade events. Samples are relabeled as the state of the previous event. Finally, samples initially indicated as missing are labeled as noise (including blinks) and event descriptives are computed (e.g., fixation duration).  
The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages \texttt{signal} [@Ligges2015] to compute velocity and acceleration signals, \texttt{depmixS4} [@Visser2010] for the HMM, \texttt{CircStats} [@Lund2018] for the von Mises distribution, and \texttt{BB} [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization. The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).

@@

## Results
### Parameter Variation
In the first part of the simulation, we examined how varying the parameters [^2] in the HMM affected the deviation of estimated parameters and accuracy of estimated state sequences. For the two-state HMM, the recovery of parameters and states was nearly perfect (all RMdSPDs < 0.1, intercepts and slopes of regression lines almost zero and one, respectively, and Cohen's kappa close to 1). We therefore chose to include the respective figures in the supplementary material.  
For the HMM with three states, the RMdSPD is shown in Figure \@ref(fig:plot-rmdspd-1-3). When response parameters (other than $a_{i=j}$) were manipulated, the RMdSPDs for $a_{12}$ and $a_{31}$ were consistently between 0.1 and 0.5. Varying $\kappa$ in states two and three led to RMdSPDs between 0.1 and 0.5 in the respective states, which we interpreted as moderate recovery. Otherwise, RMdSPDs were consistently lower than 0.1, indicating good recovery. Inspecting the regression lines between true and estimated parameters (see Figures \@ref(fig:plot-regtr-1-3) and \@ref(fig:plot-regresp-1-3)) revealed strong and unbiased linear relationships (intercepts close to zero and slopes close to one). In contrast to the two-state HMM, larger deviations and more outliers were observed. Cohen's kappa values are presented in Figure \@ref(fig:plot-acc-1-3). For most estimated models, the kappa values between true and estimated state sequences were above 0.95, meaning almost perfect agreement. However, for some models, we observed kappas clustered around zero or -0.33, which suggests that state labels were switched (see supplementary material).  
[^2]: Note that the initial state probability $\rho_i$ has $\textrm{RMdSPD} = 1$. Since the HMM only simulated one state sequence, this parameter is always either zero or one (leading to $\textrm{RMdSPD} = 1$). Therefore, we decided to exclude it from the analysis.

@@

Concerning overall disagreement, Figure \@ref(fig:plot-disag) shows that gazeHMM had less disagreement to the human coders across all events for the dynamic conditions. For the static condition, the difference to most other algorithms was only slight ($\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Image"]][3] - A2017.disag[["Image"]][-c(1:4)]))`$%), but for the dynamic conditions, the differences were substantial (video: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Video"]][4] - A2017.disag[["Video"]][-c(1:4)]))`$%) and large (dots: $\text{Med} (\Delta)  = `r abs(median(A2017.disag[["Dots"]][4] - A2017.disag[["Dots"]][-c(1:4)]))`$%).  
To explore which events gazeHMM classified differently than the average of human coder, we looked at the confusion matrix between the two (see Table \@ref(tab:tab-conf-4)). It can be seen that gazeHMM classified many fixations as smooth pursuits and vice versa. Moreover, it confused many PSOs with saccades. The heuristic to detect blinks seemed to work successfully since gazeHMM classified most blink samples in agreement with human coding and only a minor part was mistaken for saccades. Inspecting an example of gaze data classified by gazeHMM compared to human coding leads to a similar notion: Figure \@ref(fig:plot-example-classification) illustrates that gazeHMM is rapidly switching between classifying fixations and smooth pursuits, whereas the human coder treats the same data as one large smooth pursuit event. In the example, the human coder also delayed the start of the PSO compared to gazeHMM.

@@

In this report, we presented gazeHMM, a novel algorithm for classifying gaze data into eye movement events. The algorithm models velocity, acceleration, and sample-to-sample angle signals with gamma distributions and a mixture of von Mises and a uniform distribution. An HMM serves as the generative model of the algorithm and classifies the gaze samples into fixations, saccades, and optionally PSOs, and/or smooth pursuits. We showed in a simulation study that the generative model of gazeHMM recovered parameters and hidden state sequences well with a few exceptions: Adding a fourth event (i.e., smooth pursuit) to the model and noise to the generated data were the most critical factors for decreasing recovery. Furthermore, we applied gazeHMM with different numbers of states to benchmark data by @Andersson2017 and compared the model fit. The model comparison revealed that a five-state HMM had consistently most likely generated the data. This result opposed our expectation that a three-state model would be preferred for static and a four-state model for dynamic data. When comparing gazeHMM against other algorithms, gazeHMM showed mostly good agreement to human coding. On one hand, it outperformed the other algorithms in the overall disagreement to human coding for dynamic data. On the other hand, gazeHMM confused a lot of fixations with smooth pursuits, which led to a rapid switching between the two events. It also tended to mistake PSO samples as belonging to saccades.  

@@

*The submodels of gazeHMM were misspecified*. @Pohle2017 identified two scenarios in which model recovery using the ICL did not give optimal results: Outliers in the data and inadequate distributions in the response models. Both situations could apply to gazeHMM and eye movement data: Outliers occur frequently in eye-tracking data due to measurement error. Choosing adequate response distributions in HMMs is usually difficult and can depend on the individual and task from which the data are obtained from [@Langrock2015]. Moreover, gazeHMM only estimated intercepts for all parameters and thus, no time-varying covariates were included [cf. @Li2017]. This aspect could indeed oversimplify the complex nature of eye movement data.  

@@

## Conclusion
Both the simulation and validation studies showed that gazeHMM is a suitable algorithm for classifying eye movement events. For smooth pursuits, the classification is not optimal and thus not yet recommended. On one side, the algorithm has some advantages over concurrent event classification algorithms, not relying on human-labeled input being the most important one. On the other side, it is not able to perform model comparisons as expected and still poses difficulties regarding good model specifications for different eye movement events. These difficulties highlight challenges for HMMs in general but also specifically for eye movements.
